### 2022-06-21 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@be5dbd63ba5ee717e5879239f1b41feb1bd1e6e7 🚀

We downloaded the data on bats hosts of \emph{Betacoronavirus} assembled
by\textsuperscript{21} from
\texttt{https://www.viralemergence.org/betacov} on Apr.~2022, and
filtered it to ``known'' hosts (established before the emergence of
SARS-CoV-2) and ``novel'' hosts (confirmed through sampling and
competence assays since the initial data collection). The original
database was assembled by a combination of data mining and literature
surveys, including automated alerts on the ``bats'' and ``coronavirus''
keywords to identify novel empirical evidence of bats-betacoronaviruses
associations; this yielded a total of 126 known hosts, 47 of which were
novel hosts.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{63} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{64,65} in that this allows
multiple independent host-virus coevolutionary dynamics to take place in
the same location. In the colorimetric space, this correspond to yellow
-- because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{see \emph{e.g.} 66} we measured the risk
level by calculating the angular distance of the hue of each pixel to a
reference value of 60, and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{67} v1.4.0 (Algorithm FFT-NS-2, Scoring matrix
200PAM / k=2, gap open penalty 1.53m offset value 0.123) and a maximum
likelihood tree reconstructed in IQ-TREE\textsuperscript{68} v1.6.12
with ModelFinder\textsuperscript{69} ultrafast bootstrap
approximation\textsuperscript{70} with a general time reversible model
with empirical base frequencies and the 5-discrete-rate-category
FreeRaye model of nucleotide substitution (GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{71} for each of the viruses in the tree,
then averaged these at the bat species level, projected these values
onto their geographic distributions, and averaged across every bat found
in a given pixel. As such, this can be thought of as a map of the mean
evolutionary distinctiveness of the known viral community believed to be
associated with a particular subset of bats present.

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from,\textsuperscript{72,73} who
proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronaviruses tree by using the
cophenetic function in \texttt{ape};\textsuperscript{74} subsequently,
we take a principal components analysis of that distance matrix (readily
interchangeable for NMDS in this case) to project the viral tree into an
n-dimensional space. We then take the first two principal components
and, as with the evolutionary distinctiveness analysis, aggregated these
to a mean host value and projected them using a four-color bivariate
map.

@@

\leavevmode\hypertarget{ref-Cavender-Bares2009MerCom}{}%
\CSLLeftMargin{64. }
\CSLRightInline{Cavender-Bares, J., Kozak, K. H., Fine, P. V. A. \&
Kembel, S. W. The merging of community ecology and phylogenetic biology.
\emph{Ecol. Lett.} \textbf{12}, 693--715 (2009).}

\leavevmode\hypertarget{ref-Gomulkiewicz2000HotSpo}{}%
\CSLLeftMargin{65. }
\CSLRightInline{Gomulkiewicz, R., Thompson, J. N., Holt, R. D., Nuismer,
S. L. \& Hochberg, M. E. Hot Spots, Cold Spots, and the Geographic
Mosaic Theory of Coevolution. \emph{The American Naturalist}
\textbf{156}, 156--174 (2000).}

\leavevmode\hypertarget{ref-Keke2010StuSki}{}%
\CSLLeftMargin{66. }
\CSLRightInline{Keke, S., Peng, Z. \& Guohui, L. Study on skin color
image segmentation used by Fuzzy-c-means arithmetic. in \emph{2010
Seventh International Conference on Fuzzy Systems and Knowledge
Discovery} vol. 2 612--615 (2010).}

\leavevmode\hypertarget{ref-Katoh2013MafMul}{}%
\CSLLeftMargin{67. }
\CSLRightInline{Katoh, K. \& Standley, D. M. MAFFT Multiple Sequence
Alignment Software Version 7: Improvements in Performance and Usability.
\emph{Molecular Biology and Evolution} \textbf{30}, 772--780 (2013).}

\leavevmode\hypertarget{ref-Nguyen2015IqtFas}{}%
\CSLLeftMargin{68. }
\CSLRightInline{Nguyen, L.-T., Schmidt, H. A., von Haeseler, A. \& Minh,
B. Q. IQ-TREE: A Fast and Effective Stochastic Algorithm for Estimating
Maximum-Likelihood Phylogenies. \emph{Molecular Biology and Evolution}
\textbf{32}, 268--274 (2015).}

\leavevmode\hypertarget{ref-Kalyaanamoorthy2017ModFas}{}%
\CSLLeftMargin{69. }
\CSLRightInline{Kalyaanamoorthy, S., Minh, B. Q., Wong, T. K. F., von
Haeseler, A. \& Jermiin, L. S. ModelFinder: Fast model selection for
accurate phylogenetic estimates. \emph{Nature Methods} \textbf{14},
587--589 (2017).}

\leavevmode\hypertarget{ref-Hoang2018UfbImp}{}%
\CSLLeftMargin{70. }
\CSLRightInline{Hoang, D. T., Chernomor, O., von Haeseler, A., Minh, B.
Q. \& Vinh, L. S. UFBoot2: Improving the Ultrafast Bootstrap
Approximation. \emph{Molecular Biology and Evolution} \textbf{35},
518--522 (2018).}

\leavevmode\hypertarget{ref-Isaac2007MamEdg}{}%
\CSLLeftMargin{71. }
\CSLRightInline{Isaac, N. J. B., Turvey, S. T., Collen, B., Waterman, C.
\& Baillie, J. E. M. Mammals on the EDGE: Conservation Priorities Based
on Threat and Phylogeny. \emph{PLOS ONE} \textbf{2}, e296 (2007).}

\leavevmode\hypertarget{ref-Kreft2007GloPat}{}%
\CSLLeftMargin{72. }
\CSLRightInline{Kreft, H. \& Jetz, W. Global patterns and determinants
of vascular plant diversity. \emph{Proceedings of the National Academy
of Sciences} \textbf{104}, 5925--5930 (2007).}

\leavevmode\hypertarget{ref-Kreft2010FraDel}{}%
\CSLLeftMargin{73. }
\CSLRightInline{Kreft, H. \& Jetz, W. A framework for delineating
biogeographical regions based on species distributions. \emph{Journal of
Biogeography} \textbf{37}, 2029--2053 (2010).}

\leavevmode\hypertarget{ref-Paradis2019ApeEnv}{}%
\CSLLeftMargin{74. }
\CSLRightInline{Paradis, E. \& Schliep, K. Ape 5.0: An environment for
modern phylogenetics and evolutionary analyses in R.
\emph{Bioinformatics} \textbf{35}, 526--528 (2019).}

>>>>

We downloaded the data on bats hosts of \emph{Betacoronavirus} from
\texttt{https://www.viralemergence.org/betacov} on
Apr.~2022,\textsuperscript{21} and filtered it to ``known'' hosts
(established before the emergence of SARS-CoV-2) and ``novel'' hosts
(confirmed through sampling and competence assays since the initial data
collection). The original database was assembled by a combination of
data mining and literature surveys, including automated alerts on the
``bats'' and ``coronavirus'' keywords to identify novel empirical
evidence of bats-betacoronaviruses associations; this yielded a total of
126 known hosts, 47 of which were novel hosts.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{63} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{64} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{65} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{66} v1.4.0 (Algorithm FFT-NS-2, Scoring matrix
200PAM / k=2, gap open penalty 1.53m offset value 0.123) and a maximum
likelihood tree reconstructed in IQ-TREE\textsuperscript{67} v1.6.12
with ModelFinder\textsuperscript{68} ultrafast bootstrap
approximation\textsuperscript{69} with a general time reversible model
with empirical base frequencies and the 5-discrete-rate-category
FreeRaye model of nucleotide substitution (GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{70} for each of the viruses in the tree,
then averaged these at the bat species level, projected these values
onto their geographic distributions, and averaged across every bat found
in a given pixel. As such, this can be thought of as a map of the mean
evolutionary distinctiveness of the known viral community believed to be
associated with a particular subset of bats present.

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from,\textsuperscript{71,72} who
proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronaviruses tree by using the
cophenetic function in \texttt{ape};\textsuperscript{73} subsequently,
we take a principal components analysis of that distance matrix (readily
interchangeable for NMDS in this case) to project the viral tree into an
n-dimensional space. We then take the first two principal components
and, as with the evolutionary distinctiveness analysis, aggregated these
to a mean host value and projected them using a four-color bivariate
map.

@@

\leavevmode\hypertarget{ref-Gomulkiewicz2000HotSpo}{}%
\CSLLeftMargin{64. }
\CSLRightInline{Gomulkiewicz, R., Thompson, J. N., Holt, R. D., Nuismer,
S. L. \& Hochberg, M. E. Hot Spots, Cold Spots, and the Geographic
Mosaic Theory of Coevolution. \emph{The American Naturalist}
\textbf{156}, 156--174 (2000).}

\leavevmode\hypertarget{ref-Keke2010StuSki}{}%
\CSLLeftMargin{65. }
\CSLRightInline{Keke, S., Peng, Z. \& Guohui, L. Study on skin color
image segmentation used by Fuzzy-c-means arithmetic. in \emph{2010
Seventh International Conference on Fuzzy Systems and Knowledge
Discovery} vol. 2 612--615 (2010).}

\leavevmode\hypertarget{ref-Katoh2013MafMul}{}%
\CSLLeftMargin{66. }
\CSLRightInline{Katoh, K. \& Standley, D. M. MAFFT Multiple Sequence
Alignment Software Version 7: Improvements in Performance and Usability.
\emph{Molecular Biology and Evolution} \textbf{30}, 772--780 (2013).}

\leavevmode\hypertarget{ref-Nguyen2015IqtFas}{}%
\CSLLeftMargin{67. }
\CSLRightInline{Nguyen, L.-T., Schmidt, H. A., von Haeseler, A. \& Minh,
B. Q. IQ-TREE: A Fast and Effective Stochastic Algorithm for Estimating
Maximum-Likelihood Phylogenies. \emph{Molecular Biology and Evolution}
\textbf{32}, 268--274 (2015).}

\leavevmode\hypertarget{ref-Kalyaanamoorthy2017ModFas}{}%
\CSLLeftMargin{68. }
\CSLRightInline{Kalyaanamoorthy, S., Minh, B. Q., Wong, T. K. F., von
Haeseler, A. \& Jermiin, L. S. ModelFinder: Fast model selection for
accurate phylogenetic estimates. \emph{Nature Methods} \textbf{14},
587--589 (2017).}

\leavevmode\hypertarget{ref-Hoang2018UfbImp}{}%
\CSLLeftMargin{69. }
\CSLRightInline{Hoang, D. T., Chernomor, O., von Haeseler, A., Minh, B.
Q. \& Vinh, L. S. UFBoot2: Improving the Ultrafast Bootstrap
Approximation. \emph{Molecular Biology and Evolution} \textbf{35},
518--522 (2018).}

\leavevmode\hypertarget{ref-Isaac2007MamEdg}{}%
\CSLLeftMargin{70. }
\CSLRightInline{Isaac, N. J. B., Turvey, S. T., Collen, B., Waterman, C.
\& Baillie, J. E. M. Mammals on the EDGE: Conservation Priorities Based
on Threat and Phylogeny. \emph{PLOS ONE} \textbf{2}, e296 (2007).}

\leavevmode\hypertarget{ref-Kreft2007GloPat}{}%
\CSLLeftMargin{71. }
\CSLRightInline{Kreft, H. \& Jetz, W. Global patterns and determinants
of vascular plant diversity. \emph{Proceedings of the National Academy
of Sciences} \textbf{104}, 5925--5930 (2007).}

\leavevmode\hypertarget{ref-Kreft2010FraDel}{}%
\CSLLeftMargin{72. }
\CSLRightInline{Kreft, H. \& Jetz, W. A framework for delineating
biogeographical regions based on species distributions. \emph{Journal of
Biogeography} \textbf{37}, 2029--2053 (2010).}

\leavevmode\hypertarget{ref-Paradis2019ApeEnv}{}%
\CSLLeftMargin{73. }
\CSLRightInline{Paradis, E. \& Schliep, K. Ape 5.0: An environment for
modern phylogenetics and evolutionary analyses in R.
\emph{Bioinformatics} \textbf{35}, 526--528 (2019).}
